%!TEX root = /Users/gilesb/UofC/thesis/phd-thesis/phd-thesis.tex

\chapter{Quantum computation}\label{chap:quantum_computation}



\section{Quantum computation overview} % (fold)
\label{sec:quantum_computation_overview}

Quantum computation proceeds via the application of reversible transformations --- Unitary
transformations.

The semantics of quantum computation can be defined as a $\dagger$-compact closed category as
introduced in \cite{abramsky02:traces,abramsky05:abstracttraces} and completely positive maps as
discussed in \cite{selinger05:dagger}.

\begin{definition}[Dagger Category]
  A \emph{Dagger Category}\cite{selinger05:dagger} is a category \C together with an operation
  $\dagger$ that is an involutive, identity on objects, contra-variant endofunctor on \C.
\end{definition}
Recalling first that a \emph{symmetric monoidal category} is a category \B with a bi-functor $\*$,
an object $I$ and natural isomorphisms:
\begin{align*}
  a_{A,B,C}&: (A\*B)\*C \to A\* (B\*C)\\
  c_{A,B} &:A\*B \to B\*A\\
  ul_A &:A \to I \* A
\end{align*}
with standard coherence conditions, as in \cite{maclan97:categorieswrkmath}. Note that we
also have a map $ur_A: A \to A\*I$ given by $ur_A = ul_A c_{I,A}$ Furthermore, a \emph{compact
closed category} \C is a symmetric monoidal category where each object $A$ has a dual $A^{*}$
together with the maps:
\begin{align*}
  \eta_A:I \to \dual{A} \* A\\
  \epsilon_A : A\*\dual{A} \to I
\end{align*}
such that
\[
  \xymatrix@C+8pt@R-10pt{
    A \ar[r]^{ur_A} \ar@{=}[dddrr]
      & A\* I \ar[r]^(.4){A\*\eta_A}
      & A\* (\dual{A}\*A) \ar[d]^{a^{-1}}\\
    & & (A\*\dual{A})\*A \ar[d]^{\epsilon\*A}\\
    & & I\* A \ar[d]^{ul^{-1}}\\
    & & A
  }
  \quad\text{ and  }\quad
  \xymatrix@C+8pt@R-10pt{
    \dual{A} \ar[r]^{ul_{\dual{A}}} \ar@{=}[dddrr]
      & I\*\dual{A} \ar[r]^(.4){\eta_{\dual{A}\*\dual{A}}}
      & (\dual{A}\*A)\*\dual{A} \ar[d]^{a}\\
    & & \dual{A}\*(A\*\dual{A}) \ar[d]^{\dual{A}\*\epsilon}\\
    & & \dual{A}\*I \ar[d]^{ur^{-1}}\\
    & & A
  }
\]

From the above, we can define a \emph{Dagger symmetric monoidal category} and a \emph{Dagger
compact closed category}. The latter is referred to as a \emph{strongly compact closed category} in
\cite{abramsky02:traces}, where they were initially introduced. In each case, the $\dagger$ functor
is added in a way that retains coherence with the bi-functor $\*$ and with the dualizing operator.
The coherence implies that the $\dgr{i} = i^{-1}$ for the SMC isomorphisms, that $\dgr{(f\*g)} =
\dgr{f}\*\dgr{g}$ for all maps $f,g$ in the symmetric monoidal category and that
\[
  \xymatrix{
    I \ar[dr]_{\eta_A} \ar[r]^{\dgr{\epsilon_A}} & A\* \dual{A} \ar[d]^{c} \\
    &\dual{A}\*A
  }
\]
commutes for all objects $A$ in the compact closed category.


\begin{example}[\rel]
  \rel is a dagger compact closed category with the dual of an object $A$ is $A$, $\*$ is the
  cartesian product and for $R:A\to B$, we have $\dual{R} = \dgr{R} = \{(y,x) | (x,y) \in R\}$.
\end{example}
\begin{example}[\fdh]
  The category of finite dimensional Hilbert spaces, \fdh is a dagger compact closed category with
  the dual of an object $H$ is the normal Hilbert space dual $H^{*}$, the space of continuous
  linear functions from $H$ to the base field. $\*$ is the normal Hilbert space tensor and and for
  $f:A\to B$, we have $\dgr{f}$ is the unique map such that $\langle f x | y \rangle = \langle y |
  \dgr{f}x \rangle$ for all $x\in A$, $y \in B$.
\end{example}

Additionally, if one has a dagger compact closed category with biproducts where the biproducts and
dagger interact such that $\dgr{p_i} = q_i$, this is called a \emph{biproduct dagger compact closed
category}.

In \cite{selinger05:dagger}, the author continues from this point: Starting with a biproduct dagger
compact closed category $\C$, he creates a new category, $\text{CPM}(\C)$ which has the same
objects as $\C$, but morphisms $f:A \to B$ in $\text{CPM}(\C)$ are given by maps $f:\dual{A} \* A
\to \dual{B} \* B$ in $\C$ which are \emph{completely positive}. Note that \rel and \fdh are
biproduct dagger compact closed categories.

From this, the category $\text{CPM}(\C)^{\+}$, the free biproduct completion of $\text{CPM}(\C)$ is
formed, which is suitable for describing quantum computation semantics. For example, given \fdh as
our starting point, the tensor unit $I$ is the field of complex numbers. The type of
$\mathbf{qubit}$ (in \fdh and by lifting, in $\text{CPM}(\fdh)^{\+}$) is given as $I\+I$. At this
stage, the necessity of the CPM construction to model physical reality can be seen in the following
as in \fdh, the morphisms initialization of a qubit: $init:I\+I \to \mathbf{qubit}$ and destructive
measure: $meas: \mathbf{qubit} \to I\+I$ are inverses. However, in $\text{CPM}(\fdh)^{\+}$, these
same maps are given as
\[
  \dual{\mathbf{qubit}} \* \mathbf{qubit} \xrightarrow{meas} I\+
    I \xrightarrow{init}\dual{\mathbf{qubit}} \* \mathbf{qubit}
\]
by the formulae:
\[
  meas
  \begin{pmatrix}
    a & b \\
    c & d
  \end{pmatrix}
  = (a,d), \qquad init(a,d) =
  \begin{pmatrix}
    a &0 \\
    0 & d
  \end{pmatrix}.
\]
Therefore, the maps are not inverses and reflect the physical reality.


\subsection{Density matrix representation}\label{sec:density}
An alternate representation of quantum states, both pure and mixed, is via \emph{density matrices}.
If the state of a system is represented by some column vector $u$, then the matrix $u u^{*}$ is its
density matrix. Note that if $u = \nu v$ for some complex scalar $\nu$ with norm 1, then $u u^{*} =
(\nu v) (\nu v)^{*} = \nu \bar{\nu} v v^{*} = v v^{*} $. For the mixed state $\sum
\nu_{i}\{v_{i}\}$, the density matrix is $\sum \nu_{i}v_{i}v_{i}^{*}$. Density matrices are
positive hermitian matrices with trace $\le 1$. Note that the trace of the density matrix is the
probability the system has reached this particular value in the computation.

The result of applying the unitary transform $U$ to a state $u$ represented by the density matrix
$A$ is $UAU^{*}$. The measurement operation on a density matrix is derived from the measurement
effects on the \qubit. For example, consider the density matrix for $q=\alpha\kz+\beta\ko$,
$\begin{pmatrix}\alpha\bar{\alpha}&\alpha\bar{\beta}\\ \beta\bar{\alpha} &
\beta\bar{\beta}\end{pmatrix}$. Measuring this \qubit gives either
$\begin{pmatrix}\alpha\bar{\alpha}&0\\ 0& 0\end{pmatrix}$ with probability $|\alpha|^{2}$ or
$\begin{pmatrix}0&0\\ 0 & \beta\bar{\beta}\end{pmatrix}$ with probability $|\beta|^{2}$. If the
results of the measurment are not used, this will result in the density matrix
$\begin{pmatrix}\alpha\bar{\alpha}&0\\ 0 & \beta\bar{\beta}\end{pmatrix}$. This extends linearly so
that if a \qubit is measured in the system whose density matrix is \qsmat{A}{B}{C}{D}, the result
will be the mixed density matrix \qsmat{A}{0}{0}{D}.

It is possible to create a complete partial order on density matrices.

\begin{definition}[L\"owner partial order]\label{def:lownerorder}
  For square complex matrices $A,B$ of the same size, define $A \le B$ if $B-A$ is positive.
\end{definition}

\begin{lemma} \label{lemma:cpodensity}
  Designate $D_{n}$ to be the density matrices of size $n\times n$, then the poset $(D_{n}, \le)$
  is a complete partial order.
\end{lemma}
\begin{proof}
  See \cite{selinger04:qpl}, pp 13--14.
\end{proof}

% section quantum_computation_overview (end)
\include{chapters/section/dagger-categories}
\include{chapters/section/quantum-semantics}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../phd-thesis"
%%% End:
