% textmate project ignores: !(/\.(?!htaccess)[^/]*|\.(tmproj|o|pyc|toc|aux|xyc|bbl|blg|lof|lot|out|synctex\.gz)|/Icon\r|/svn-commit(\.[2-9])?\.tmp)$

%   Modified   92-09-18      Add references to dissertation        D. Teale
%                            Add approval page to toc
%                            Add reference to Title Degree on approval page
%   Modified  2006-09-12     Added geometry package to set up UofC thesis margins
%                            Removed includeprompt option N. Mancell
%
% Modified  2012-04-15  T.Zhang
% 1, the ``fancy'' package is canceled. The page style is changed from ``myheadings''
%    and ``headings'' to ``plain'' to move the page number (footer) from top right to bottom center.
% 2, The parameter  of using ``geometry'' package is changed from four different values
%     (top=1in, bottom=1.22in, left=1.40in, right=0.850in) to one single value:
%     1in (top=1in, bottom= 1in, left= 1in, right= 1in)
% 3, ``List of Figures'' is modified to ``List of Figures and Illustrations''
% 4, A new page ``List of Symbols, Abbreviations and Nomenclature'' is created.
%    Its page number is following ``List of Figures and Illustrations'' with Roman numerals.
%    A separate file ``symbols.tex'' is created for students to put content into the list.





\documentclass{ucalgthes1}
\usepackage{etex}
\usepackage[letterpaper,top=1in, bottom= 1in, left= 1in, right= 1in]{geometry}
%Define other usepackages here

\input{packageuses}

\graphicspath{{./diagrams/}{.}}
%
% captionskips do not seem to be defined in uc class
\makeatletter
%%\newlength\abovecaptionskip
%%\newlength\belowcaptionskip
\setlength\abovecaptionskip{10\p@}
\setlength\belowcaptionskip{0\p@}
\makeatother
\input{Qcircuit}
\input{phd-macros}
\newcommand{\thesistitle}{An investigation of some theoretical aspects of reversible computing}
\title{\thesistitle \\
\bigskip  }
%
%            Insert the correct information between the {}
%
\author{Brett Gordon Giles}
\thesisyear{2014}
\thesis{dissertation}    % the word dissertation can be inserted between {}
\monthname{September}
\dept{COMPUTER SCIENCE}
\degree{DOCTOR OF PHILOSOPHY}
%
%                    End of supplied information
%
\begin{document}
\makethesistitle
\pagenumbering{roman}     % resets page counter to one
\setcounter{page}{1}
\chapter*{UNIVERSITY OF CALGARY \\ FACULTY OF GRADUATE STUDIES}
\thispagestyle{empty}
The undersigned certify that they have read, and recommend
to the Faculty of Graduate Studies for acceptance, a \Thesis\ entitled
``\thesistitle'' submitted by \Author\
in partial fulfillment of the requirements for the degree of
\Degree.\\
\begin{signing}{Department of Computer Science}
\signline
Dr.~J.R.B.~Cockett\\
Professor, Department of Computer Science \\
Thesis Supervisor\\
\signline
Dr. Kristine Bauer \\
Assistant Professor, Department of Mathematics \\
Supervisory committee member  \\
\signline
Dr.~Peter~Høyer \\
Associate Professor, Department of Computer Science \\
Supervisory committee member  \\
\newsigncolumn
\signline
Dr.~Gilad~Gour \\
Associate Professor, Applied Mathematics\\
Internal external examiner  \\
%\newsigncolumn         use this command to start a new column if necessary
\signline
Dr. Bob Coecke\\
Professor of Quantum Foundations, Logics and Structures\\
Department of Computer Science,  Oxford University \\
External Examiner
\end{signing}
%
\newpage
\phantomsection
\altchapter[Abstract]{\bf{Abstract}}
The categorical semantics of reversible computing must be a category which combines the concepts of
partiality and the ability to reverse any map in the category. Inverse categories, restriction
categories in which each map is a partial isomorphism, provide exactly this structure. This thesis
explores inverse categories and relates them to both quantum computing and standard non-reversible
computing. The former is achieved by showing that commutative Frobenius algebras form an inverse
category. The latter is by establishing the equivalence of the category of discrete inverse
categories to the category of discrete Cartesian restriction categories --- this is the main result
of this thesis. This allows one to transfer the formulation of computability given by Turing
categories onto discrete inverse categories.

\newpage
\phantomsection
\altchapter[Acknowledgements]{\bf{Acknowledgements}}

First and foremost, many thanks to my supervisor of the past many, many years, Dr. Robin
Cockett. His indefatigable thirst for mathematical exploration of computer science has been an
inspiration for both my master's degree and my doctoral studies.

Secondly, thank you to Dr. Peter Selinger, both for his paper on the semantics of quantum computing
which inspired me to write LQPL and the opportunity to work for him at MSRI in Berkeley for the fall
of 2012. His clarity of exposition in research writing remains a target to which I aspire.

Thank you to Dr. Kristine Bauer and Dr. Peter Høyer for serving on my supervisory committee,
partaking in my qualifying exam and serving on my thesis defence committee. I always felt your doors
were open to me.

Thank you to Dr. Gilad Gour for being my internal examiner.

Thank you to Dr. Bob Coecke for agreeing to be my external examiner.

Thanks to my fellow graduate students (and post-docs) over the years for great conversations, people
to hang with while at conferences and critiques of research. Thanks to Craig Pastro, Xiuzhan Guo,
Dana Harrington, Ross Duncan, Jon Grattage, Sean Nichols, Ning Tang, Alexander Green, Neil Ross,
Jonathan Gallagher, Chad Nester and Prashant Kumar.

Finally, and most of all, thank you so much to all of my family. My wife, for being with me for all
these years and proof-reading both my master's and this thesis --- learning more about math than she
ever thought she would; my kids, who still remember me; and my grandkids --- who likely don't remember a time
when Grandpa was not in school.
\newpage
\phantomsection
\altchapter[Dedication]{\bf{Dedication}}
\vspace{3in}
\begin{flushright}
To my wife, Marie Gelinas Giles.
\end{flushright}


\begin{singlespace}
\newpage
\phantomsection
\tableofcontents
\pagestyle{plain}
\newpage
\phantomsection
\listoftables
\pagestyle{plain}
\newpage
\phantomsection
\listoffigures
\pagestyle{plain}
\clearpage
\clearpage          % otherwise tables will be numbered wrong
\end{singlespace}
\newpage
\phantomsection
\chapter*{\bf{List of Symbols, Abbreviations and Nomenclature}\hfill} \addcontentsline{toc}{chapter}{List of Symbols}
\listofsymbols
%\pagestyle{plain}
%\clearpage


\pagenumbering{arabic}
\input{chapters/introduction}
\input{chapters/introduction-to-categories}
\input{chapters/restriction-categories}
\input{chapters/inverse-categories}
\input{chapters/inverse-cats-completion}
\input{chapters/coproducts_and_disjointness}
\input{chapters/disjoint-sums-and-tensors}
\input{chapters/alternate-matrix-categories}
\input{chapters/distributive-inverse-categories}
\input{chapters/commutative-frobenius}
\input{chapters/turing-categories}
\input{chapters/conclusion}

\begin{singlespace}
\bibliography{bibentries-phd}
\bibliographystyle{plain}
\end{singlespace}
\appendix
%\input{appendix1}
\end{document}
