%!TEX root = /Users/gilesb/UofC/thesis/phd-thesis/phd-thesis.tex
\chapter{Disjointness in Inverse Categories} % (fold)
\label{cha:disjointness_in_inverse_categories}

In this chapter we will explore coproduct like structure in inverse categories. We start by showing,
that similar to the product, having coproducts is too restrictive a notion for inverse categories: An
inverse category with a coproduct is a preorder.
Nonetheless it is possible to define coproduct like structures in an inverse category. To introduce
this structure we define a ``disjointness'' relation between parallel maps of an inverse category
and whence a ``disjoint join'' for disjoint maps. We will then show how a tensor product satisfying
some specific conditions gives rise to both a disjointness relation and a disjoint join. Such a
tensor product provides the proposed replacement for ``coproducts'' in an inverse category.



\section{Coproducts in inverse categories} % (fold)
\label{sec:coproducts_in_inverse_categories}

A restriction category can have coproducts and an initial object. For example, \Par (sets and
partial functions) has coproducts.

\begin{definition}\label{def:restriction_coproduct}
  In a restriction category \R, a coproduct is a \emph{restriction coproduct} when the embeddings
  $\cpa$ and $\cpb$ are total.
\end{definition}
\begin{lemma}\label{lem:restriction_coproduct_is_restriction_functor}
  A restriction coproduct in $\R$ satisfies:
  \begin{enumerate}[{(}i{)}]
    \item $\restr{f+g} = \restr{f} + \restr{g}$ which means $+$ is a restriction functor.
    \item $\nabla:A+A\to A$ is total.
    \item $?:0 \to A$ is total, where $0$ is the initial object in the restriction category.
  \end{enumerate}
\end{lemma}
\begin{proof}
  \prepprooflist
  \begin{enumerate}[{(}i{)}]
    \item Totality of $\cpa$ and $\cpb$.
    \item \textbf{$+$ is a restriction functor.}
      Consider the diagram:
      \[
        \xymatrix @C=30pt @R=35pt{
          A \ar[rr]^{f} \ar[dr]_{\cpa} && A' \ar[dr]^{\cpa'}\\
          &A+B \ar@{.>}[rr]^{f+g} & & A' + B'\\
          B \ar[rr]_{g} \ar[ur]^{\cpb} && B'. \ar[ur]_{\cpb'}
        }
      \]
      In order to show $\restr{f+g} = \restr{f} + \restr{g}$, it suffices to show that
      $\cpa\restr{f+g} = \cpa(\restr{f} + \restr{g}) = \restr{f}\cpa$.
      \begin{align*}
        \cpa\restr{f+g} &= \restr{\cpa(f+g)}\cpa & \rfour\\
        &= \restr{f \cpa'} \cpa &\text{coproduct diagram}\\
        &= \restr{f \restr{\cpa'}} \cpa
          & \text{Lemma }\refitem{lem:restrictionvarious}{lemitem:rv_3}\\
        &= \restr{f} \cpa & \cpa'\text{ total}.
      \end{align*}
    \item   \textbf{$\nabla:A+A\to A$ is total.}
      By the definition of $\nabla$ ($=\<1 | 1 \>$) and the co-product, the following diagram
      commutes,
      \[
        \xymatrix @C=30pt @R=35pt{
          &A + A \ar[d]_{\nabla}\\
          A \ar@{=}[r] \ar[ur]^{\cpa}&A \ar@{=}[r] & A \ar[ul]_{\cpb}
        }
      \]
      resulting in:
      \[
         \cpa \restr{\nabla}  = \restr{\cpa \nabla} \cpa = \restr{1} \cpa = \cpa.
      \]
      Similarly, $\cpb \restr{\nabla} = \cpb$, hence, the restriction of $\nabla$ is $1$ and
      therefore $\nabla$ is total.
    \item  \textbf{$\why:0 \to A$ is total.}
      This follows from
      \[
        \xymatrix @C=30pt @R=35pt{
          0 \ar[dr]_{\why} \ar[r]^{\cpb} &A + 0 \\
          & A \ar@{=}[u]
        }
      \]
      so $\why$ can be defined as the total coproduct injection.
  \end{enumerate}



\end{proof}

Recall that when an object is both initial and terminal, it is referred to as a zero object and
denoted as $\zeroob$. This gives rise to the zero map $\zeroob_{A,B}: A \to \zeroob \to B$ between any two objects.
Note that for all $f:D\to A, g:B\to C$, we have $f\zeroob_{A,B} g = \zeroob_{D,C}$ as this factors through
the zero object.

\begin{definition}\label{def:restriction_zero}
  Given a restriction category \X with a zero object, then $\zeroob$ is a \emph{restriction zero} when
  for each object $A$ in \X we have $\restr{\zeroob_{A,A}} = \zeroob_{A,A}$.
\end{definition}

\begin{lemma}[Cockett-Lack\cite{cockettlack2004:restcategories3}, Lemma 2.7]\label{lem:restriction_zero_equivalences}
  For a restriction category \X, the following are equivalent:
  \begin{enumerate}[{(}i{)}]
    \item \X has a restriction zero;
    \item \X has an initial object $0$ and terminal object $1$ and each initial map $z_A$ is a
      restriction monic;
    \item \X has a terminal object $1$ and each terminal map $t_A$ is a restriction retraction.
  \end{enumerate}
\end{lemma}
% subsection coproducts (end)




\subsection{Inverse categories with restriction coproducts} % (fold)
\label{sub:inverse_categories_with_restriction_coproducts}


\begin{proposition}\label{prop:inverse_category_with_coproducts_is_pre-order}
  An inverse category \X with restriction coproducts is a pre-order.
\end{proposition}
\begin{proof}
  By Lemma~\ref{lem:restriction_coproduct_is_restriction_functor}, we know $\nabla$ is total and
  therefore $\nabla\inv{\nabla} = 1$. From the coproduct diagrams, we have $\cpa\nabla =1$ and
  $\cpb\nabla = 1$. But this gives us $\inv{\nabla}\inv{\cpa} = \inv{(\cpa\nabla)} = 1$ and
  similarly $\inv{\nabla}\inv{\cpb} =1$. Hence, $\inv{\nabla} = \cpa$ \emph{and} $\inv{\nabla} =
  \cpb$.

  This means for parallel maps $f,g:A \to B$, we have
  \[
    f = \cpa \<f|g\> = \inv{\nabla} \<f|g\> = \cpb\<f|g\> = g
  \]
  and therefore \X is a pre-order.
\end{proof}
% subsection inverse_categories_with_restriction_coproducts (end)
% section coproducts_in_restriction_categories (end)

\section{Disjointness in an inverse category} % (fold)
\label{sec:disjointness_in_an_inverse_category}
In the following, we will add a relation, \emph{disjointness}, and a construction, \emph{disjoint
  join}, on parallel maps in  an inverse category with a restriction
zero and zero maps. The disjoint join is evocative of the join as defined in
Section~\ref{sub:joins_in_restriction_categories}. We will begin by defining disjointness on
parallel maps and then show this is equivalent to a definition of disjointness on the restriction
idempotents.

From this point forward in the thesis, we will work with a number of relations and operations on
parallel pairs of maps. Suppose we have a relation $\lozenge$ between maps $f,g:B\to C$, i.e., $f
\lozenge g$. We will refer to $\lozenge$ as \emph{stable} whenever given $h:A \to B$, then $h f
\lozenge h g$. We will refer to $\lozenge$ as \emph{universal} whenever given $k:C \to D$, then $f
k \lozenge g k$.


\begin{definition}\label{def:disjointness_relation}
  In an inverse category \X with zero maps, the relation $\cdperp$ between two parallel
  maps $f, g:A \to B$ is called a \emph{disjointness relation} when it satisfies the following
  properties:
  \begin{align*}
    \axiom{Dis}{1}\quad &\text{ For all }f:A\to B,\ f\cdperp 0;&\text{(Zero is disjoint to all maps)}\\
    \axiom{Dis}{2}\quad &f\cdperp g\text{ implies }\rst{f} g = 0; &\text{(Disjoint maps have no intersection)}\\
    \axiom{Dis}{3}\quad &f\cdperp g,\ f' \le f,\ g' \le g\text{ implies }f' \cdperp
    g';&\text{(Disjointness is down closed)}\\
    \axiom{Dis}{4}\quad &f\cdperp g\text{ implies }g \cdperp f; &\text{(Symmetric)}\\
    \axiom{Dis}{5}\quad &f\cdperp g\text{ implies }h f \cdperp h g; &\text{(Stable)}\\
    \axiom{Dis}{6}\quad &f\cdperp g\text{ implies }\rst{f} \cdperp \rst{g}
      \text{ and }\rg{f}\cdperp\rg{g}; &\text{(Closed under range and restriction)}\\
    \axiom{Dis}{7}\quad &\rst{f}\cdperp \rst{g},\ \rg{h}\cdperp \rg{k}\text{ implies }f h
      \cdperp g k &\text{(Determined by restriction/range).}
  \end{align*}
  When $f\cdperp g$, we will say $f$ is \emph{disjoint} from $g$.
\end{definition}

\begin{lemma}\label{lem:disjointness_equivalent_axioms}
  In \ref{def:disjointness_relation}, given $\axiom{Dis}{1-5}$, we may
  replace $\axiom{Dis}{6}$ and $\axiom{Dis}{7}$ by:
  \begin{align*}
    \axiom{Dis}{6'}\quad &f\cdperp g\text{ if and only if }\rst{f} \cdperp \rst{g}
      \text{ and }\rg{f}\cdperp\rg{g}.
  \end{align*}
\end{lemma}
\begin{proof}
  Given $\axiom{Dis}{6}$ and $\axiom{Dis}{7}$, the \emph{only if} direction of $\axiom{Dis}{6'}$ is
  immediate. To show the \emph{if} direction, assume $\rst{f} \cdperp \rst{g}$ and
  $\rg{f}\cdperp\rg{g}$. This also means that $\rst{\rst{f}} \cdperp \rst{\rst{g}}$. Then, by
  \axiom{Dis}{7}, $\rst{f} f \cdperp \rst{g}g$ and therefore $f \cdperp g$, showing $\axiom{Dis}{6}$
  and $\axiom{Dis}{7}$ imply $\axiom{Dis}{6'}$.

  Conversely, assume we are given \axiom{Dis}{6'}. Then, \axiom{Dis}{6} follows immediately. To show
  \axiom{Dis}{7}, assume we have $\rst{f}\cdperp \rst{g},\ \rg{h}\cdperp \rg{k}$. As
  $\rst{f h} \le \rst{f}$ and $\rst{g k}\le\rst{g}$, by \axiom{Dis}{3}, we know that
  $\rst{f h} \cdperp \rst{g k}$. Similarly, $\wrg{f h} \le \rg{h}$ and $\wrg{g k} \le \rg{k}$,
  giving us $\wrg{f h}\cdperp \wrg{g k}$. Then, from \axiom{Dis}{6'} we may conclude
  $f h \cdperp g k$, showing \axiom{Dis}{7} holds.
\end{proof}
\begin{lemma}\label{lem:disjointness_various}
  In an inverse category \X with $\cdperp$ a disjointness relation:
  \begin{enumerate}[{(}i{)}]
    \item $f \cdperp g$ if and only if $\inv{f}\cdperp \inv{g}$; \label{lemitem:djv_inverses}
    \item $f \cdperp g$ implies $f h \cdperp g h$ (Universal);\label{lemitem:djv_universal}
    \item $f \cdperp g$ implies $f\rg{g} = 0$; \label{lemitem:djv_disjoint_composition_is_0}
    \item if $m,n$ are monic, then $f m \cdperp g n$ implies $\rst{f} \perp \rst{g}$;
      \label{lemitem:djv_monic_implies}
    \item if $m,n$ are monic, then $\inv{m} f \cdperp \inv{n}g$ implies $\rg{f} \perp \rg{g}$.
      \label{lemitem:djv_inv_monic_implies}
  \end{enumerate}
\end{lemma}
\begin{proof}
  \prepprooflist
  \begin{enumerate}[{(}i{)}]
    \item Assume $f \cdperp g$. By \axiom{Dis}{6}, we have $\rst{f} \cdperp \rst{g}$ and
      $\rg{f}\cdperp\rg{g}$. Since $\rg{f} = \rst{\inv{f}}$ and $\rst{f} = \wrg{\inv{f}}$, this
      means $\rst{\inv{f}} \cdperp \rst{\inv{g}}$ and $\wrg{\inv{f}}\cdperp\wrg{\inv{g}}$. By
      \axiom{Dis}{6'} from Lemma~\ref{lem:disjointness_equivalent_axioms}, we have
      $\inv{f} \cdperp \inv{g}$. The converse follows with  a similar argument.
    \item Assume $f \cdperp g$. By \ref{lemitem:djv_inverses}, we have $\inv{f}\cdperp\inv{g}$. By
      \axiom{Dis}{5}, $\inv{h}\inv{f}\cdperp\inv{h}\inv{g}$, giving us $\inv{(f h)} \cdperp
      \inv{(g h)}$. Applying \ref{lemitem:djv_inverses}, we now have $f h \cdperp g h$.
    \item Assume $f \cdperp g$. From \ref{lemitem:djv_inverses} and reflexivity, we know that
      $\inv{g} \cdperp \inv{f}$ and therefore $\rst{\inv{g}}\inv{f} = \rg{g}\inv{f}= 0$. However, in
      an inverse category, $\inv{0} = 0$ and therefore $0 = \inv{(\rg{g}\inv{f})} = f \inv{\rg{g}} =
      f \rg{g}$.
    \item Assume $f m \cdperp g n$ where $m, n$ are monic. By \axiom{Dis}{6}, this gives us
      $\rst{f m} \cdperp \rst{g n}$. By Lemma~\ref{lem:restrictionvarious},
      $\rst{f m} = \rst{f \rst{m}} = \rst{f 1} = \rst{f}$ and therefore $\rst{f} \cdperp \rst{g}$.
    \item By assumption, we have $\inv{m} f \cdperp  \inv{n}g$ and therefore $\inv{f}m \cdperp
      \inv{g} n$. By \ref{lemitem:djv_monic_implies}, this means $\rst{\inv{f}} \cdperp
      \rst{\inv{g}}$ and hence $\rg{f} \cdperp \rg{g}$.
  \end{enumerate}
\end{proof}

We may equivalently define a disjointness relation by a relation on the restriction idempotents, \open{A}.

\begin{definition}\label{def:disjointness_in_open_x}
  Given an inverse category \X, a relation $\ocdperpsub{A} \subseteq \open{A}\times\open{A}$ for
  each $A \in \mathrm{ob}(\X)$, is an \emph{open disjointness} relation when for all $e, e' \in
  \open{A}$
  \begin{align*}
    \axiom{$\mathcal{O}$dis}{1}\quad &1 \ocdperpsub{A} 0; &\text{(Zero disjoint to all idempotents)}\\
    \axiom{$\mathcal{O}$dis}{2}\quad &e \ocdperpsub{A} e' \text{ implies }e' \ocdperpsub{A} e;
    &\text{(Symmetric)}\\
    \axiom{$\mathcal{O}$dis}{3}\quad &e \ocdperpsub{A} e' \text{ implies }e e' = 0; &\text{(Disjoint implies no intersection)}\\
    \axiom{$\mathcal{O}$dis}{4}\quad &e \ocdperpsub{A} e' \text{ implies }\rst{f e} \ocdperpsub{B}
      \rst{f e'}\text{ for all }f:B \to A; &\text{(Stable)}\\
    \axiom{$\mathcal{O}$dis}{5}\quad &e \ocdperpsub{A} e' \text{ implies }\wrg{e g} \ocdperpsub{C}
      \wrg{e' g}\text{ for all }g:A \to C; &\text{(Universal)}\\
    \axiom{$\mathcal{O}$dis}{6}\quad &e \ocdperpsub{A} e',\ e_1 \le e,\ e_1' \le e'
      \text{ implies }e_1 \ocdperpsub{A} e_1' &\text{(Down closed)}.
  \end{align*}
\end{definition}

We will normally write $\ocdperp$ rather than $\ocdperpsub{A}$ where the object is clear.

\begin{proposition}\label{prop:disjointness_is_open_disjointness}
  If $\cdperp$ is a disjointness relation in $\X$, then $\ocdperp = \cdperp \intersection
  (\union_{A\in\X} \open{A}\times \open{A})$, the restriction of $\cdperp$ to the restriction
  idempotents, is an open disjointness relation.
\end{proposition}
\begin{proof}
  \prepprooflist
  \setlist[enumerate,1]{leftmargin=1.75cm}
  \begin{enumerate}
    \item[\axiom{$\mathcal{O}$dis}{1}] This follows immediately from \axiom{Dis}{1} by taking
      $f = 1$.
    \item[\axiom{$\mathcal{O}$dis}{2}] Reflexivity follows directly from  \axiom{Dis}{4}.
    \item[\axiom{$\mathcal{O}$dis}{3}] By \axiom{Dis}{2}, $0 = \rst{e} e' = e e'$.
    \item[\axiom{$\mathcal{O}$dis}{4}] Given $e \cdperp e'$, we have $f e \cdperp f e'$ by
      \axiom{Dis}{5}. Then, by \axiom{Dis}{6} we may conclude $\rst{f e} \cdperp \rst{f e'}$.
    \item[\axiom{$\mathcal{O}$dis}{5}] This follows from the above item, using $\inv{g}$ for $f$.
      This means we have $\rst{\inv{g}e} \cdperp \rst{\inv{g}e'}$. But this gives us $\rst{\inv{(e
      g)}} \cdperp \rst{\inv{(e' g)}}$. Recalling from
      Lemma~\ref{lem:inverse_categories_are_range_categories} that $\rg{k} = \rst{\inv{k}}$, we may
      conclude $\wrg{e g} \cdperp \wrg{e' g}$.
    \item[\axiom{$\mathcal{O}$dis}{6}] Assuming $e \cdperp e'$ and $e_1 \le e,\ e'_1 \le e'$, by
      \axiom{Dis}{3}, $e_1 \cdperp e'_1$.
  \end{enumerate}
  Therefore, $\ocdperp = \cdperp \intersection (\union_{A\in\X} \open{A}\times \open{A})$ acts as an
  open disjointness relation on $\open{A}^2$.

\end{proof}

For the converse, if $\ocdperp$ is an open disjointness relation in $\X$, then we may define a relation
$\perpab \subseteq \union_{A,B\in\X} \X(A,B)\times \X(A,B)$ on parallel maps by
\[
  f\perpab g \iff  \rst{f}\ocdperp\rst{g} \text{ and } \rg{f}\ocdperp\rg{g}.
\]
Note the use of the $\rg{f}\ocdperp\rg{g}$. By \rrone, we know $\rst{\rg{f}} = \rg{f}$ and
therefore it is a restriction idempotent in $\open{B}$.

We then have:

\begin{proposition}\label{prop:extended_disjointness_is_a_disjointness_relation}
  If $\ocdperp$ is an open disjointness relation in \X, then
  \[
    f\perpab g \iff  \rst{f}\ocdperp\rst{g}  \text{ and }  \rg{f}\ocdperp\rg{g}
  \]
  is a disjointness relation in \X.
\end{proposition}
\begin{proof}
  \prepprooflist
  \setlist[enumerate,1]{leftmargin=1.5cm}
  \begin{enumerate}
    \item[\axiom{Dis}{1}] We need to show $f \cdperp 0$ for any $f$. We know that $1 \ocdperp 0$
      and therefore $\rst{f} \ocdperp 0$ and $\rg{f} \ocdperp 0$, as $\rst{f}\le 1$ and $\rg{f}\le
      1$. This gives us $f \cdperp 0$.
    \item[\axiom{Dis}{2}] Assume $f \cdperp g$, i.e., $\rst{f}\ocdperp \rst{g}$. Then, $\rst{f}g =
      \rst{f}\rst{g}g = 0 g = 0$.
    \item[\axiom{Dis}{3}] We are given $f \cdperp g$, $f' \le f$ and $g' \le g$. By
      Lemma~\refitem{lem:restriction_cats_are_partial_order_enriched}{lemitem:rst_ordering_2}
      $\rst{f'} \le \rst{f}$ and $\rst{g'} \le \rst{g}$. Then, by \axiom{$\mathcal{O}$dis}{6}, as
      $\rst{f} \ocdperp \rst{g}$ we have $\rst{f'} \ocdperp \rst{g'}$. By
      Lemma~\refitem{lem:ordering_of_restriction_and_range}{lemitem:ordering_2}, we have $\wrg{f'}
      \le \rg{f}$ and $\wrg{g'} \le \rg{g}$. Then, by \axiom{$\mathcal{O}$dis}{6}, as $\rg{f}
      \ocdperp \rg{g}$ we have $\wrg{f'} \ocdperp \wrg{g'}$. This means $f' \cdperp g'$.
    \item[\axiom{Dis}{4}] Reflexivity of $\cdperp$ follows immediately from the reflexivity of
      $\ocdperp$.
    \item[\axiom{Dis}{5}] Assume $f \cdperp g$, i.e., $\rst{f}\ocdperp \rst{g}$ and $\rg{f}
      \ocdperp \rg{g}$. Then we have $\rst{hf}\ocdperp \rst{h g}$ by \axiom{$\mathcal{O}$dis}{4}. By
      Lemma~\refitem{lem:ordering_of_restriction_and_range}{lemitem:ordering_1} we have $\wrg{h f}
      \le \rg{f}$ and $\wrg{h g} \le \rg{g}$. Therefore we have $\wrg{h f}\ocdperp \wrg{h g}$ by
      \axiom{$\mathcal{O}$dis}{6} and therefore $h f \cdperp h g$.
    \item[\axiom{Dis}{6}] Assume $f\perpab g$. Then both $\rst{f}\ocdperp\rst{g}$ and
      $\rg{f}\ocdperp\rg{g}$. By Lemma~\ref{lem:basic_range_category_properties} we know for any map
      $h$, $\rg{\rst{h}} = \rst{h}$ and by Lemma~\ref{lem:restrictionvarious} we have $\rst{\rst{h}}
      = \rst{h}$. Thus, we have both $\rst{\rst{f}} \ocdperp \rst{\rst{g}}$ and $\rg{\rst{f}}
      \ocdperp \rg{\rst{g}}$ and therefore $\rst{f} \perpaa \rst{g}$. Similarly for any map $h$,
      \rrone gives $\rst{\rg{h}} = \rg{h}$ and Lemma~\ref{lem:basic_range_category_properties} gives
      $\rg{\rg{h}} = \rg{h}$. Thus, we have $\rst{\rg{f}} \ocdperp \rst{\rg{g}}$ and $\rg{\rg{f}}
      \ocdperp \rg{\rg{g}}$ giving us $\rg{f} \perpvv{B}{B} \rg{g}$.
    \item[\axiom{Dis}{7}] We assume $\rst{f} \cdperp \rst{g}$ and $\rg{h} \cdperp \rg{k}$. Then,
      we have $\rst{f} \ocdperp \rst{g}$ and $\rg{h} \ocdperp \rg{k}$. By
      Lemma~\refitem{lem:restriction_cats_are_partial_order_enriched}{lemitem:rst_ordering_3}, we
      have $\rst{f h} \le \rst{f}$ and $\rst{g k} \le \rst{g}$. Therefore, $\rst{f h} \ocdperp
      \rst{g k}$ by \axiom{$\mathcal{O}$dis}{6}. By
      Lemma~\refitem{lem:ordering_of_restriction_and_range}{lemitem:ordering_1}, $\wrg{f h} \le
      \rg{h}$ and $\wrg{g k} \le \rg{k}$, giving us $\wrg{f h}\ocdperp \wrg{g k}$ also by
      \axiom{$\mathcal{O}$dis}{6}. This means $f h \cdperp g k$.
  \end{enumerate}
\end{proof}
\begin{theorem}\label{thm:open_disjointness_is_disjointness}
   To give a disjointness relation $\perp$ on $\X$ is to give an open disjointness relation
   $\ocdperp$ on \X.
\end{theorem}
\begin{proof}
  We must show there is a bijection between disjointness relations and open disjointness
  relations. That is, give a disjointness relation $\perp$, we know we may generate an open
  disjointness relation, $\ocdperp$, from it. We then need to show that the disjointness relation
  generated from $\ocdperp$ is in fact $\perp$.

  Suppose we are given the disjointness relation $\perp$. By
  Proposition~\ref{prop:disjointness_is_open_disjointness}, this is an open disjointness $\ocdperp =
  \perp \intersection (\union_{A\in\X}\open{A}\times\open{A})$.

  By Proposition~\ref{prop:extended_disjointness_is_a_disjointness_relation}, $\perpab$ defined by
  \[
    f\perpab g \iff  \rst{f}\ocdperp\rst{g}  \text{ and }  \rg{f}\ocdperp\rg{g}.
  \]
  is a disjointness relation on $\X$.

  Assume $f\perp g$. We have $\rst{f}\ocdperp\rst{g}$ and $\rg{f}\ocdperp\rg{g}$ by
  \axiom{Dis}{6} and Proposition~\ref{prop:disjointness_is_open_disjointness}. Then by its
  definition, we have $f \perpab g$.

  Assume $f \perpab g$. This means we must have had $\rst{f}\ocdperp\rst{g}$ and
  $\rg{f}\ocdperp\rg{g}$. Therefore $\rst{f}\perp\rst{g}$ and  $\rg{f}\perp\rg{g}$. By
  Proposition~\ref{lem:disjointness_various}, we have $f \perp g$.

  We have shown $f \perp g \iff f\perpab g$. We may also conclude that if we started with an open
  disjointness relation $\ocdperp$ and used it to construct $\perpab$, then the relation $\perpab
  \intersection (\union_{A\in\X}\open{A}\times\open{A})$ would again be $\ocdperp$.

  Hence we have shown a bijection between disjointness relations and open disjointness relations on
  an inverse category $\X$.


  % following used to be part of longer murky proof - should not bee needed...

  % Now, suppose we are given the open disjointness relation $\ocdperp$. Similar to above, we can
  % construct the extended disjointness relation $\perp$ by
  % Definition~\ref{def:extended_open_disjointness}. From the disjointness relation $\perp$, we have
  % the open disjointness relation $\overline{\perp}$ by
  % Lemma~\ref{prop:disjointness_is_open_disjointness}.

  % Assume $e \ocdperp e'$. As this means both $\rst{e} \ocdperp \rst{e'}$ and
  % $\rg{e} \ocdperp \wrg{e'}$, we have $e \perp e'$. By
  % Proposition~\ref{prop:disjointness_is_open_disjointness} this means $e \overline{\perp} e'$.

  % If we are given that $e \overline{\perp} e'$, then we know that $e \perp e'$ by
  % Proposition~\ref{prop:disjointness_is_open_disjointness}. From
  % Definition~\ref{def:extended_open_disjointness}, this requires that $\rst{e} \ocdperp \rst{e'}$
  % and $\rg{e} \ocdperp \wrg{e'}$, but that just means $e \ocdperp e'$.
\end{proof}


Disjointness is additional structure on a restriction category, i.e., it is possible to have more
than one disjointness relation on the category. To see this, consider the trivial disjointness
relation, where $f \perp_0 g$ if and only if $f = 0$ or $g = 0$. As $\rst{0} = 0 = \rg{0}$, $f \le 0
\iff f = 0$ and $h 0 =0$ for any map $h$, axioms \axiom{Dis}{1} through \axiom{Dis}{7} are
immediately satisfied. $\perp_0$ is definable on any inverse category with zero maps, hence is
definable on \pinj. However, \pinj does have a more useful disjointness relation:
\begin{example}[\pinj has a disjointness relation]\label{ex:pinj_has_a_disjointness_relation}
  Consider the inverse category \pinj, introduced in Example~\ref{ex:pinj_is_a_restriction_category}
  and Example~\ref{ex:pinj_is_a_discrete_inverse_category}.
  We note the restriction zero is the empty set, $\emptyset$, with initial and terminal maps being
  $\emptyset$ and therefore $\zeroob_{A,B} = \emptyset$.

  We may define the disjointness relation $\perp$ by $f \perp g$ if and only if
  $\rst{f}\intersection\rst{g}=\emptyset$ and $\rg{f}\intersection\rg{g} = \emptyset$. It is then
  reasonably straightforward to verify \axiom{Dis}{1} through \axiom{Dis}{7}. For example, take
  \axiom{Dis}{7}:
  \begin{proof}
    We are given $\rst{f}\cdperp\rst{g}$ and $\rg{h}\cdperp\rg{k}$. This means
    \[
      \rst{f}\intersection\rst{g}=\emptyset \text{ and }\rg{h}\intersection\rg{k} = \emptyset.
    \]
    As discussed in Example~\ref{ex:pinj_has_meets}, we have
    $\rst{m n} \subseteq \rst{m}$ and that $\wrg{m n} \subseteq \rg{n}$. Hence we have
    \begin{align*}
      \rst{f h}\intersection\rst{g k}&\subseteq \rst{f}\intersection \rst{g}= \emptyset\\
      \wrg{f h}\intersection\wrg{g k}&\subseteq \rg{h}\intersection \rg{k}= \emptyset.
    \end{align*}
    Therefore, $f h \cdperp g k$.
  \end{proof}

  We see that we may define $f \perp_0 g$ on \pinj as well: $f \perp_0 g$ if and only if one of $f$
  or $g$ is the restriction $\zeroob$, $\emptyset$. As $\zeroob = \rst{\zeroob} = \rg{\zeroob} = h
  \zeroob = \zeroob k$, all of the seven disjointness axioms are easily verifiable.
\end{example}

Although disjointness is additional structure on a restriction category, one can use the
disjointness structure of a base category (or categories) to define a disjointness structure
on the product category.

\begin{lemma}\label{lem:disjointness_is_derivable_on_a_product_category}
  If $\X$ and $\Y$ are inverse categories with restriction zeros and respective disjointness
  relations $\perp$ and $\perp'$, then we may construct a disjointness relation
  $\perp_{\times}$ on $\X\times\Y$.
\end{lemma}
\begin{proof}
  The restriction, the inverse and the restriction zero on the product category are defined pointwise.
  \begin{itemize}
    \item If $(f,g)$ is a map in $\X\times\Y$, then $\inv{(f,g)} = (\inv{f}, \inv{g})$;
    \item If $(f,g)$ is a map in $\X\times\Y$, then $\rst{(f,g)} = (\rst{f}, \rst{g})$;
    \item The map $(\zeroob_X,\zeroob_Y)$ is the restriction zero in $\X\times\Y$.
  \end{itemize}

  Following this pattern, for $(f,g)$ and $(h,k)$ maps in $\X\times\Y$, $(f,g) \perp_{\times}(h,k)$
  iff $f\perp h$ and $g\perp' k$.

  Verifying the disjointness axioms is straightforward, we show axioms 2 and 5. Proofs of the
  others are similar.
  {
  \setlist[itemize,1]{leftmargin=1.5cm}
  \begin{itemize}
    \item [\axiom{Dis}{2}]: Given $(f,g)\perp_{\times}(h,k)$, we have
      $\rst{(f,g)}(h,k) =  (\rst{f},\rst{g})(h,k) = (\rst{f} h, \rst{g} k) = (\zeroob, \zeroob) = \zeroob$.
    \item [\axiom{Dis}{5}]: We are given $(f,g)\perp_{\times}(h,k)$. Consider the map $z = (x,y)$
      in $\X \times \Y$. We know that $x f \perp x h$ and $y g \perp y k$, therefore we have
      $z(f,g) = (x f, y g) \perp_{\times} (x h, y k) = z(h,k)$.
  \end{itemize}
  }
\end{proof}
% subsection disjointness_relations (end)

\section{Disjoint joins} % (fold)
\label{sec:disjoint_joins}

We now consider additional structure on the inverse category, dependent upon the disjointness
relation. Note that while we have worked with binary disjointness up to this point, one may extend
the concept to sets of maps simply by considering disjointness pairwise:
$\cdperp \{f_1,f_2,\ldots,f_n\} \definedas f_i \cdperp f_k$ whenever $i \ne j$.

\begin{definition}\label{def:disjoint_join}
  An inverse category with $\X$ with a restriction $0$ and a disjointness relation $\perp$ has
  \emph{disjoint joins} when there is a binary operator on disjoint parallel maps:
  \[
    \infer{f \djoin g : A \to B}{f: A\to B,\ g: A\to B,\ f \perp g}
  \]
  such that the following hold:
  \begin{align*}
    \axiom{DJ}{1}\quad & f \le f \djoin g\text{ and }g \le f \djoin g; \\
    \axiom{DJ}{2}\quad & f \le h,\ g \le h\text{ and }f\perp g\text{ implies }f \djoin g \le h;\\
    \axiom{DJ}{3}\quad & h(f \djoin g) = h f \djoin h g. \text{ (Stable)} \\
    \axiom{DJ}{4}\quad & \cdperp \{f, g, h\}\text{ if and only if }f \perp (g\djoin h).\\
  \end{align*}
  The binary operator, $\djoin$, is called the \emph{disjoint join}.
\end{definition}

Given a specific disjointness relation on a category, there is only one disjoint join:
\begin{lemma}\label{lem:disjoint_join_is_unique}
  Suppose \X in an inverse category with a disjointness relation $\perp$, then if  $\djoin$ and
  $\altjoin$ are disjoint joins for $\perp$ then $\djoin = \altjoin$.
\end{lemma}
\begin{proof}
  \axiom{DJ}{1} gives us:
  \[
    f,g \le f\djoin g\text{ and }f,g \le f \altjoin g.
  \]
  By \axiom{DJ}{2}, we may therefore conclude $f \djoin g \le f \altjoin g$ and
  $f \altjoin g \le f\djoin g$, hence $f \djoin g = f \altjoin g$.
\end{proof}
\begin{lemma}\label{lem:join_is_associative_and_commutative_monoid}
  In an inverse category with disjoint joins, the disjoint join respects the restriction and is
  universal. Additionally, it is a partial associative and
  commutative operation, with identity $\zeroob$. That is, the following hold:
  \begin{enumerate}[{(}i{)}]
    \item $\rst{f\djoin g} = \rst{f} \djoin \rst{g}$;
    \item $(f \djoin g)k = f k \djoin g k$ (Universal);
    \item $f \perp g,\ g\perp h,\ f\perp h$ implies that $(f\djoin g)\djoin h = f\djoin(g\djoin
      h)$; \label{lemitem:associative_join}
    \item $f \perp g$ implies $f \djoin g = g \djoin f$; \label{lemitem:commutative_join}
    \item $f \djoin \zeroob = f$. \label{lemitem:identity_for_join}
  \end{enumerate}
\end{lemma}
\begin{proof}
  \prepprooflist
  \begin{enumerate}[{(}i{)}]
    \item As $\rst{f}, \rst{g} \le \rst{f\djoin g}$, we immediately have
      $\rst{f} \djoin \rst{g} \le \rst{f \djoin g}$. To show the other direction, consider
      \begin{align*}
        \rst{f}(\rst{f} \djoin \rst{g}) (f \djoin g)
        &= (\rst{f}\,\rst{f} \djoin \rst{f}\rst{g})(f \djoin g) &\text{\axiom{DJ}{3}}\\
        &= \rst{f} (f\djoin g) &\text{Lemma~\ref{lem:restrictionvarious}, \axiom{Dis}{2}}\\
        &= f.
      \end{align*}
      Hence, we have $f \le (\rst{f} \djoin \rst{g}) (f \djoin g)$ and similarly, so is $g$. By
      \axiom{DJ}{2} and that $\rst{f} \djoin \rst{g}$ is a restriction idempotent, we then have
      \[
        f \djoin g \le (\rst{f} \djoin \rst{g}) (f \djoin g) \le f \djoin g
      \]
      and therefore $f \djoin g = (\rst{f} \djoin \rst{g}) (f \djoin g)$. By
      Lemma~\ref{lem:restriction_cats_are_partial_order_enriched}, $\rst{f\djoin g} \le \rst{f}
      \djoin \rst{g}$ and so $\rst{f\djoin g} = \rst{f}\djoin \rst{g}$.
    \item First consider when $f, g$ and $k$ are restriction idempotents, say $e_0, e_1 $ and $e_2$.
      Then, we have $(e_0 \djoin e_1)e_2 = e_2(e_0 \djoin e_1) = e_2 e_0 \djoin e_2 e_1 =
      e_0 e_2 \djoin e_1 e_2$.
      Next, note that for general $f,g,h$, we have $f k \djoin g k \le (f\djoin g) k$ as both
      $f k, g k \le (f\djoin g) k$.
      By Lemma~\ref{lem:restriction_cats_are_partial_order_enriched}, we need only show that their
      restrictions are equal:
      \begin{align*}
        \rst{(f\djoin g)k} &= \rst{\rst{f\djoin g}(f\djoin g) k } & \rone \\
        & = \rst{f\djoin g}\, \rst{(f\djoin g)k } & \rthree\\
        & = (\rst{f}\djoin \rst{g}) \rst{(f\djoin g)k }  & \text{previous item}\\
        & = \rst{f}\,\rst{(f\djoin g)k } \djoin \rst{g}\, \rst{(f\djoin g)k }
          &\text{idempotent universal}\\
        & = \rst{\rst{f} (f\djoin g)k} \djoin \rst{\rst{g} (f\djoin g)k} & \rthree\\
        & = \rst{f k} \djoin \rst{g k}\\
        & = \rst{f k \djoin g k}.
      \end{align*}
      Therefore, as the restrictions are equal, we have shown $(f\djoin g)k = f k \djoin g k$.
    \item \emph{Associativity}: Note that \axiom{DJ}{4} shows that both sides of the equation
      exist. To show they are equal, we show that they are less than or equal to each other.

      From the definitions, we know that $f\djoin g, h \le (f\djoin g)\djoin h$, which also means
      $f, g \le (f\djoin g)\djoin h$. Similarly, $g\djoin h \le (f\djoin g)\djoin h $ and then $f
      \djoin (g\djoin h)\le (f\djoin g)\djoin h$. Conversely, $f,g,h \le f \djoin (g\djoin h)$ and
      therefore $(f\djoin g)\djoin h \le f \djoin (g\djoin h)$ and both sides are equal.
    \item \emph{Commutativity}: Note first that both $f$ and $g$ are less than or equal to both
      $f\djoin g$ and $g \djoin f$, by \axiom{DJ}{1}. By \axiom{DJ}{2}, we have $f \djoin g \le
      g\djoin f$ and $g\djoin f \le f \djoin g$ and we may conclude $f \djoin g = g \djoin f$.
    \item \emph{Identity}: By \axiom{DJ}{1}, $f \le f \djoin \zeroob$. As $\zeroob \le f$ and $f \le
      f$, by \axiom{DJ}{2}, $f \djoin \zeroob \le f$ and we have $f = f \djoin \zeroob$.
  \end{enumerate}
\end{proof}

Note that the previous lemma and proof of associativity allows a simple inductive argument which
shows that having binary disjoint joins extends to disjoint joins of an arbitrary
finite collection of disjoint maps.

Recalling our notation for disjointness of a set of maps,
$\djoin \{f_i\}$ will mean the disjoint join of all maps $f_i$, i.e.,
$f_1 \djoin f_2 \djoin \cdots \djoin f_n$.

\begin{lemma}\label{lem:arbitrary_disjoint_joins}
  In an inverse category $\X$ with disjoint joins,
  \begin{enumerate}[{(}i{)}]
    \item $\cdperp \{f_i\}$ if and only if  $\djoin \{f_i\}$ is
      defined,\label{lemitem:arbitrary_disjoint_joins_1}
    \item if $f_i, g_j : A \to B$ and $\cdperp_{i\in I} \{f_i\}$ and $\cdperp_{j\in J} \{g_j\}$,
      then $\djoin_{i\in I} \{f_i\} \cdperp \djoin_{j\in J} \{g_j\}$ if and only $f_i \cdperp g_j$
      for all $i\in I$ and $j\in J$. \label{lemitem:arbitrary_disjoint_joins_2}
  \end{enumerate}
\end{lemma}
\begin{proof}
  For \ref{lemitem:arbitrary_disjoint_joins_1}, using \axiom{Dj}{4}, proceed as in the proof of
  Lemma~\refitem{lem:join_is_associative_and_commutative_monoid}{lemitem:associative_join},
  inducting on $n$.

  To show \ref{lemitem:arbitrary_disjoint_joins_2}, first assume $\djoin \{f_i\} \cdperp \djoin
  \{g_j\}$. By \axiom{Dj}{4} and associativity, we have $\djoin \{f_i\} \cdperp g_j$ for each
  $j$. Using the reflexivity of $\cdperp$, \axiom{Dj}{4} and associativity, we have $f_i \cdperp
  g_j$ for each $i$ and $j$.

  Next, assume $f_i \cdperp g_j$ for each $i$ and $j$. Then by \axiom{Dj}{4} and associativity, $f_i
  \cdperp \djoin \{g_j\}$ for each $i$. Applying \axiom{Dj}{4} again, we have $\djoin \{f_i\} \cdperp
  \djoin \{g_j\}$.
\end{proof}

Clearly the product of two inverse categories with disjoint joins has a disjoint join:
\begin{lemma}\label{lem:disjoint_join_is_in_product_category}
  Given $\X, \Y$ are inverse categories with disjoint joins, $\djoin$ and $\djoin'$ respectively,
  then the category $\X \times \Y$ is an inverse category with disjoint joins.
\end{lemma}
\begin{proof}
  From Lemma~\ref{lem:disjointness_is_derivable_on_a_product_category}, we know $\X\times\Y$ has a
  disjointness relation that is defined point-wise.  We therefore define $\djoin_{\times}$ the
  disjoint join on $\X\times\Y$ by
  \begin{equation}
    (f,g)\djoin_{\times}(h,k) = (f \djoin h, g \djoin' k). \label{eq:disjoint_join_on_product}
  \end{equation}
  We now prove each of the axioms in Definition~\ref{def:disjoint_join} hold.
  \setlist[itemize,1]{leftmargin=1.5cm}
  \begin{itemize}
    \item [\axiom{DJ}{1}] From Equation~\ref{eq:disjoint_join_on_product}, we see that since
      $f,h \le f \djoin h$ and $g,k \le g \djoin' k$, we have $(f,g) \le (f,g)\djoin_{\times}(h,k)$
      and  $(h,k) \le (f,g)\djoin_{\times}(h,k)$.
    \item [\axiom{DJ}{2}] Suppose $(f,g) \le (x,y)$, $(h,k) \le (x,y)$ and $(f,g) \perp_{\times}
      (h,k)$. Then regarding it point-wise, we have $(f,g)\djoin_{\times}(h,k) = (f \djoin h, g
      \djoin' k) \le (x,y)$.
    \item [\axiom{DJ}{3}] $(x,y)\left((f,g)\djoin_{\times}(h,k)\right) = (x(f \djoin h), y(g
      \djoin' k)) = (x f \djoin x h, y g \djoin' y k) = (x f,y g)\djoin_{\times}(x h,y k) =
      ((x,y)(f,g))\djoin_{\times}((x,y)(h,k))$.
    \item [\axiom{DJ}{4}] Given $\perp_{\times}\!\![(f,g),(h,k),(x,y)]$, we know $f \perp (h\djoin
      x)$ and $g \perp' (k\djoin' y)$. Hence, $(f,g) \perp_{\times}((h,k)\djoin_{\times}(x,y))$. The
      opposite direction is similar.
  \end{itemize}
\end{proof}

\begin{example}[\pinj has a disjoint join]\label{ex:pinj_has_a_disjoint_join}
  Continuing from Example~\ref{ex:pinj_has_a_disjointness_relation}, we show that \pinj has disjoint
  joins. If $f = \{(a,b)\}$ and $g=\{(a',b'\}$ are disjoint parallel maps in \pinj from $A$ to $B$,
  define $f\djoin g \definedas \{(a'',b'') | (a'',b'') \in f\text{ or }(a'',b'') \in g\}$, i.e., the
  union of $f$ and $g$.

  This is still a partial injective map, due to the requirement of disjointness. Recall that $f\perp
  g$ means that $\rst{f}\intersection \rst{g} = \emptyset$ and $\rg{f}\intersection\rg{g} =
  \emptyset$ and that the respective meets will also be $\emptyset$. The empty meet of the
  restrictions means that $f\djoin g$ is still a partial function,  as each $a''$ will appear only
  once. The empty meet of the ranges gives us that $f\djoin g$ is injective, because each $b''$ is
  unique.

  The axioms for disjoint joins all hold:
  \begin{enumerate}
    \item[\axiom{DJ}{1}] By construction, both $f$ and $g$ are less than $f\djoin g$.
    \item[\axiom{DJ}{2}] $f \le h,\ g \le h$ means that $h$ must contain all of the $(a,b)\in f$ and
      $(a',b') \in g$ and therefore $f\djoin g \le h$.
    \item[\axiom{DJ}{3}] Suppose $h:C\to A = \{(c,\dot{a})\}$. Then
      \begin{align*}
        h(f \djoin g) &= \{(c,\dot{b}) |(\exists a,\dot{a}.\dot{a} = a, (a,\dot{b}) \in f,
        (c,\dot{a}) \in h) \\
        & \qquad\qquad \text{ or }(\exists a',\dot{a}.\dot{a} = a', (a',\dot{b}) \in g, (c,\dot{a}) \in h)\}\\
        & = \{(c,\dot{b}) |\exists a, \dot{a}.\dot{a} = a, (a,\dot{b}) \in f, (c,\dot{a}) \in h\} \bigcup\\
        & \qquad\qquad\{(c,\dot{b}) | \exists a',\dot{a}.\dot{a} = a', (a',\dot{b}) \in g, (c,\dot{a}) \in h\}\\
        &= h f \djoin h g
      \end{align*}
    \item[\axiom{DJ}{4}] Suppose $\cdperp [f, f', f'']$, $f=\{(a,b)\}, f'=\{(a',b')\},
      f''=\{(a'',b'')\}$. Then the set $\{a\}$ does not intersect either $\{a'\}$ nor $\{a''\}$ and
      similarly for the sets $\{b\}, \{b'\}$ and $\{b''\}'$. Thus we have $f \perp (g\djoin h)$. The
      reverse direction is argued similarly.
  \end{enumerate}
\end{example}
% section disjoint_joins (end)


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../phd-thesis"
%%% End:
