%!TEX root = /Users/gilesb/UofC/thesis/phd-thesis/phd-thesis.tex

\chapter{Introduction to Categories}\label{chap:introduction_to_categories}

This chapter introduces categories and fixes notation for them. More
details for category theory can be found from, e.g., \cite{barr:ctcs}, \cite{cockett2009:ctcs},
\cite{maclan97:categorieswrkmath} and \cite{various:nlab}.

\section{Definition of a category}
\label{sec:definition_of_a_category}


A category may be defined in a variety of equivalent ways. As much of our work will involve the
exploration of partial and reversible maps, we choose a definition that highlights the algebraic
nature of these.

\begin{definition}\label{def:category}
  A \emph{category} $\A$ is a directed graph consisting of objects $\objects{\A}$ and maps $\arrows{\A}$. Each $f\in
  \arrows{\A}$ has two associated objects in $\objects{\A}$, called the domain, $\catdomain{f}$, and codomain,
  $\catcodomain{f}$. When $\catdomain{f}$ is the object $X$  and $\catcodomain{f}$ is the object $Y$, we
  will write $f:X \to Y$. For $f, g \in \A_m$, if $f:X\to Y$ and $g:Y \to Z$, there is a map called
  the \emph{composite} of $f$ and $g$, written $f g$,\footnote{Note that composition is written in
    diagrammatic order throughout this thesis.} such that $f g:X \to Z$. For any $W \in \A_o$ there is
  an \emph{identity} map $1_W:W \to W$. Additionally, these two axioms must hold:
  \begin{align*}
    \catone &\text{ for }f:X \to Y,\ 1_X f = f = f 1_Y;&\text{(Unit laws)}\\
    \cattwo &\text{ given }f:X \to Y,\ g:Y \to Z\text{ and }h: Z\to W\text{, then }f (g h) = (f g) h.&\text{( Associativity)}
  \end{align*}
\end{definition}

For a category $\A$, given objects $X,Y\in\objects{A}$, the set of maps between $X$ and $Y$ is
referred to as the \emph{hom-set} of $\A$ between $X$ and $Y$ and written as $\A(X,Y)$.

Throughout   this thesis, we will be working with \emph{small} categories, that is, those categories
whose collection of maps and collection of objects is, in fact, a set. We will give categories of
``all'' sets as an example, and the reader can take that to mean all our sets are very small and all
belong to a some sufficiently large set, $\mathcal{U}$.

We give a few examples of categories:

\begin{example}\label{ex:small_finite_categories}
  We set $\mathbf{1}$ to be the category consisting of a single object $A$ and the identity arrow
  $1:A\to A$. This  is obviously a category, with $\mathbf{1}_o = A$ and
  $\catdomain{1} = \catcodomain{1} = A$.

  There is also the category $\mathbf{2}$:
  \[
    \xymatrix{
      A\ar[rr]^{f} \ar@(ul,dl)[]_{1_A}& & B  \ar@(ur,dr)[]^{1_B}
    }
  \]
  where $f$ is the only non-identity arrow.
\end{example}

\begin{example}[Preorders are categories]\label{ex:preorders-are-categories}
  Take any partially ordered set $(P,\le)$ and define $f:a \to b$ for $a,b\in P$ if and only if
  $a\le b$. This is a category as we always have:
  \begin{enumerate}[{(}i{)}]
  \item $a \le a$ (Identity);
  \item $a \le b$ and $b \le c$ implies $a\le c$ (Composition).
  \end{enumerate}
  Note that we have at most one map between any two objects in $P$, hence \catone and \cattwo are
  immediately satisfied.
\end{example}
\begin{example}[Dual Category]\label{ex:dual_category}
Given a category $\B$, we may form the \emph{dual} of $\B$, written $\dual{\B}$ as the following
category:
\category{The objects of $\B$}{$\dual{f}:B\to A$ in $\dual{\B}$ when $f:A\to B$ in $\B$}{
The identity maps of $\B$}{If $f g = h$ in $\B$, $\dual{g} \dual{f} = \dual{h}$}
\end{example}

Note the format of the previous example, where we list the four basic requirements of a
category. This is typically how we will present categories in this thesis. Depending upon the
complexity of the definition, we may add further proof that it meets \catone and \cattwo.

The previous example is an important one, as we will often speak of \emph{dualizing} a notion, or
that concept ``x'' is the \emph{dual} of concept ``y''. This means that when ``y'' holds in a
category \B, the ``x'' holds in $\dual{\B}$.

\section{Properties of maps} % (fold)
\label{sub:properties_of_maps}
Many useful properties of maps are generalizations of notions used for sets and functions. We
present a few of these in Table~\ref{tab:properties_of_maps_in_categories}, together with their
categorical definition. Throughout Table~\ref{tab:properties_of_maps_in_categories}, $e,f,g$ are
maps in a category $C$ with $e:A \to A$ and $f,g:A \to B$.
\begin{table}[!htbp]
  \begin{center}
    \begin{tabular}{|p{1in}p{1in}p{3.78in}|}
      \hline
      {\bf Sets} & {\bf Categorical Property} & {\bf Definition}\\
      \hline
      \hline
      Injective & Monic & $f$ is monic whenever $h f = k f$ means that $h = k$.\\
      & & \\
      \hline
      Surjective & Epic & {The dual notion to monic, $g$ is epic whenever $g h = g k$ means that $h = k$.
      A map that is both monic and epic is called \emph{bijic}.}\\
      & & \\
      \hline
      Left~Inverse & Section & $f$ is a section when there is a map $\categorysection{f}$ such that $f \categorysection{f} = 1_A$. $f$
      is also referred to as the \emph{left inverse} of $\categorysection{f}$.\\
      & & \\
      \hline
      Right Inverse & Retraction & $f$ is a retraction when there is a map $\retraction{f}$ such that $\retraction{f} f = 1_B$.
      $f$ is also referred to as the \emph{right inverse} of $\retraction{f}$. A map that is both a section and a
      retraction is called an \emph{isomorphism}.\\
      & & \\
      \hline
      Idempotent & Idempotent & An endomorphism $e$ is idempotent whenever $e e = e$.\\
      & & \\
      \hline
    \end{tabular}
  \end{center}
  \caption{Properties of Maps In Categories}
  \label{tab:properties_of_maps_in_categories}
\end{table}

There are number of basic properties of maps enumerated in Table~\ref{tab:properties_of_maps_in_categories}.

\begin{lemma}\label{lem:categorical_properties_of_maps}
  In a category \B,
  \begin{enumerate}[{(}i{)}]
    \item If $f,g$ are monic, then $f g$ is monic.
    \item If $f g$ is monic, then $f$ is monic.
    \item $f$ being a section means it is monic.
    \item $f, g$ sections implies that $f g$ is a section.
    \item $f g$ a section means $f$ is a section.
    \item If $f:A \to B$ is both a section and a retraction, then $\categorysection{f} = \retraction{f}$, where $\categorysection{f}$ and $\retraction{f}$ are as
      defined in Table~\ref{tab:properties_of_maps_in_categories}.
    \item  $f$ is an isomorphism if and only if it is an epic section.
  \end{enumerate}
\end{lemma}
\begin{proof}
  \prepprooflist
  \begin{enumerate}[{(}i{)}]
    \item Suppose $h f g = k f g$. As $g$ is monic, $h f = k f$. As $f$ is monic, this gives us $h =
      k$ and therefore $f g$ is monic.
    \item See \cite{barr:ctcs}, chapter 2.
    \item Suppose $h f = k f$. Then $h f \categorysection{f} = k f \categorysection{f}$ giving us $h
      1 = k 1$ and therefore $h = k$  and $f$ is monic.
    \item We are given $f\categorysection{f} =1$ and $g\categorysection{g} =1 $. But then $f g
      \categorysection{g}\categorysection{f} = f \categorysection{f} = 1$ and $f g$ is a section.
    \item We are given there is some $h$ such that $(f g) h = 1$. This means $f (g h) =1$ and $f$ is
      a section.
    \item See \cite{cockett2009:ctcs}, Lemma 1.2.2.
    \item See \cite{cockett2009:ctcs}, Lemma 1.2.3.
  \end{enumerate}

\end{proof}

Note there are corresponding properties for epics and retractions, obtained by dualizing the
statements of Lemma~\ref{lem:categorical_properties_of_maps}.

Suppose $f:A \to B$ is a retraction with left inverse $\retraction{f}:B \to A$. Note that $f
\retraction{f}$ is idempotent as $f \retraction{f} f \retraction{f} = f 1_B \retraction{f} = f
\retraction{f}$. If we are given an idempotent $e$, we say $e$ is \emph{split} if there is a
retraction $f$ with $e = f \retraction{f}$.

In general, not all idempotents in a category will split. The following construction allows us to
create a category based on the original one in which all idempotents do split.

\begin{definition}\label{def:split_category}
  Given a category $\B$ and a set of idempotents $E$ of $\B$, we may create the category of \B split
  over the idempotents $E$. This is  normally written as $\spl{E}{\B}$, and defined as:
  \categoryns{$(A,e)$, where $A$ is an object of \B, $e:A\to A$ and $e\in E$.}
    {$f_{d,e}:(A,d)\to(B,e)$ is given by $f:A\to B$ in \B, where $f = d\uts f e$.}
    {The map $e_{e,e}$ for $(A,e)$.}
    {Inherited from \B.}
  When $E$ is the set of all idempotents in $\B$, we write $\spl{}{\B}$.
\end{definition}
This is the standard idempotent splitting construction, variously known as the Karoubi
envelope (whence the notation) or Cauchy completion.

% subsection properties_of_maps (end)

\section{Functors and natural transformations} % (fold)
\label{ssub:functors_and_natural_transformations}

\begin{definition}\label{def:functor}
  A map $F:\X \to \Y$ between categories, as in Definition~\ref{def:category}, is called a
  \emph{functor}, provided it satisfies the following:
  \begin{itemize}
    \item[\axiom{F}{1}] $F(\catdomain{f}) = \catdomain{F(f)}$ and $F(\catcodomain{f}) = \catcodomain{F(f)}$;
    \item[\axiom{F}{2}] $F(f g) = F(f)F(g)$.
  \end{itemize}
\end{definition}

\begin{lemma}\label{lem:cat_is_a_category}
  Categories and functors form a category \cat.
\end{lemma}
\begin{proof}
  \prepprooflist
  \categoryns{Categories.}{Functors.}{The identity functor which takes an object to the same object
    and a map to the same map.}{Given $F:\A\to\B$, $G:\B\to\D$, define the functor  $F G:\A\to\D$ such that
    $F G(x) = G(F(x))$ which is clearly associative.}
  Note this is often regarded as a \emph{large} category, as its collection of objects is not a set.
\end{proof}


A functor $F:\B \to \D$ induces a map between hom-sets in $\B$ and hom-sets in $\D$. For
each object $A,B$ in $\B$ we have the map:
\[
  F_{AB} : \B(A,B) \to \D(F(A),F(B)).
\]

\begin{definition}\label{def:full_functor}
  A functor $F:\X\to\Y$ is \emph{full} when for each pair of objects $A,B$ in $\X$, and map
  $g:FA \to F B$ in $\Y$, there is a map $f:A\to B$ in $\X$ such that $F f = g$.
\end{definition}

\begin{definition}\label{def:faithful_functor}
  A functor $F:\X\to\Y$ is \emph{faithful} when for parallel maps $f, f'$, if $Ff = Ff'$ then $f = f'$.
\end{definition}
We may also consider the notion of containment between categories:

\begin{definition}\label{def:subcategories}
  Given the categories $\B$ and $\D$, we say that \B is a \emph{subcategory} of \D when each object
  of \B is an object of \D and when each map of \B is a map of \D.
\end{definition}

When $\B$ is a subcategory of $\D$, the functor $J:\B \to \D$ which takes each object to itself in
$\D$ and each map to itself in $\D$ is called the inclusion functor. When $J$ is a full functor, we
say $\B$ is a full subcategory of $\D$.

We now have the machinery to discus the relationship between a category \B and its splitting
$\spl{}{\B}$:

\begin{lemma}\label{lem:split_category_splits_and_has_category}
  Given a category $\B$, then it is a full subcategory of $\spl{}{\B}$ and all idempotents split
  in  $\spl{}{\B}$.
\end{lemma}
\begin{proof}
  First, recall $\spl{}{\B}$ means we are splitting over all idempotents in $\B$, including the
  identity maps.
  We identify each object $A$ in $\B$ with the object $(A,1)$ in $\spl{}{\B}$. The only maps between
  $(A,1)$ and $(B,1)$ in $\spl{}{\B}$ are the maps between $A$ and $B$ in $\B$, hence we have a
  full subcategory.

  Suppose we have the map $d_{e,e}: (A,e) \to (A,e)$ with $d d = d$, i.e., it is idempotent in $\B$
  and $\spl{}{\B}$. In $\spl{}{\B}$, we have the maps $d_{e,d}:(A,e) \to (A,d)$  as
  $e d d = e (e d  e) d = e d e d = d d = d$ and
  $d_{d,e}:(A,d) \to (A,e)$ as $d d e = d e = (e d e) e = e d e = d$. The compositions
  are $d_{d,e} d_{e,d} = d_{d,d} = 1_{(A,d)}$ and $d_{e,d} d_{d,e} = d_{e,e}$. Hence,
  it is a splitting of the map  $d_{e,e}$.
\end{proof}


Functors with two arguments, e.g., $F:\A \times \A \to \A$ which satisfy \axiom{F}{1} and
\axiom{F}{2} for each argument independently are called \emph{bi-functors}.


We will often restrict ourselves to specific classes of functors which either \emph{preserve} or
\emph{reflect} certain characteristics of the domain category or codomain category.

\begin{definition}\label{def:diagram_in_a_category}
  Given a category $\cS$, a \emph{diagram}
  in a category $\B$ of \emph{shape} $\cS$ is a functor $D:\cS \to \B$.
\end{definition}

\begin{definition}\label{def:property_of_a_diagram}
  A \emph{property} of a diagram $D$, written $P(D)$, is a logical relation expressed using the
  objects and maps of the diagram $D$.
\end{definition}

For example,   $P(f:A \to B) \definedas \exists h : B \to A. h f = 1_A$ expresses that $f$ is a retraction.

\begin{definition}\label{def:functor_preserving_and_reflecting_a_property}
  A functor $F$ \emph{preserves} the property $P$ over maps $f_i$ and objects $A_j$ when:
  \[
     P(f_1,\ldots,f_n, A_1,\ldots,A_m) \implies P(F(f_1),\ldots,F(f_n), F(A_1),\ldots,F(A_m)).
  \]
  A functor $F$ \emph{reflects} the property $P$ over maps $f_i$ and objects $A_j$ when:
  \[
    P(F(f_1),\ldots,F(f_n), F(A_1),\ldots,F(A_m)) \implies P(f_1,\ldots,f_n, A_1,\ldots,A_m).
  \]
\end{definition}

For example, all functors preserve the properties of being an idempotent or a retraction or section,
but in general, not the property of being monic.

\begin{definition}\label{def:natural_transformation}
  Given functors $F,G:\X \to \Y$, a \emph{natural transformation} $\alpha:F \natto G$ is a collection
  of maps in $\Y$, $\alpha_X : F(X) \to G(X)$, indexed by the objects of $\X$ such that for all
  $f:X_1 \to X_2$ in $\X$ the following diagram in $\Y$ commutes:
  \[\xymatrix @R+10pt @C+10pt{
      F(X_1) \ar[r]^{F(f)} \ar[d]_{\alpha_{X_1}} & F(X_2) \ar[d]^{\alpha_{X_2}}\\
      G(X_1) \ar[r]_{G(f)} &  G(X_2).
    }
  \]
\end{definition}

In the case where a natural transformation is a collection of isomorphisms, we write $\alpha: F
\cong G$ or simply $F\cong G$.

% subsection functors_and_natural_transformations (end)


\section{Adjoint functors and equivalences}
\label{sec:adjoint-functors-and-equivalences}

Referring to Section~\ref{ssub:functors_and_natural_transformations}, we consider relationships
between categories. While two categories may be isomorphic (i.e. there is an invertible functor
between them), category theory normally considers two alternate relations between categories:
Equivalence and adjointness.

Adjointness refers to an \emph{adjoint pair} of functors. There are various equivalent ways of
defining adjoints and we will approach them using universality.

\begin{definition}\label{def:universal_pair}
  Given $G:\B\to\A$ is a functor and $A\in \A$, then an object $U$ in \B together with a map
  $\eta_A:A\to G(U)$ is a \emph{universal pair} for the $G$ at $A$ if whenever there is a map
  $f:A\to G(Y)$, there is a unique $f^{\#}:U\to Y$ such that
  \[
    \xymatrix@C+15pt{
      A \ar[r]^{\eta_A} \ar[dr]_{f} & G(U) \ar@{.>}[d]^{G(f^{\#})}\\
      & G(Y)
    }
  \]
  is a commutative diagram.
\end{definition}

Consider what happens when we have the situation above, but the $U$ in $\B$ is definable
based on the $A\in\A$.
\begin{lemma}\label{lem:adjoint-setup}
  Suppose $G:\B\to\A$ is a functor such that for each $A\in\objects{\A}$, there is an map on objects
  $F:\objects{\A}\to\objects{\B}$ such that  $(F(A),\eta_A)$ is a universal pair. Then:
  \begin{itemize}
    \item $F$ is a functor with $F(g) = (g\eta)^{\#}$;
    \item $\eta_A: A \to G(F(A))$ is a natural transformation;
    \item $\epsilon_B = 1_{G(B)}^{\#}:F(G(B)) \to B$ is a natural transformation;
    \item The triangle equalities, $\eta_{G(B)}G(\epsilon_B) = 1_{G(B)}$ and $F(\eta_A)\epsilon_{F(A)} =
      1_{F(A)}$ hold.
  \end{itemize}
  Conversely, if we have functors $F,G$ with transformation $\eta$ and $\epsilon$ which satisfy the
  triangle identities, then each $(F(A),\eta_A)$ is universal for $G$ at $A$.
\end{lemma}
\begin{proof}
  See Proposition 2.2.2\cite{cockett2009:ctcs}.
\end{proof}

When we have this occurring we say $F$ is \emph{left adjoint} to $G$ (and $G$ is the
\emph{right adjoint} of $F$). The transformation $\eta$ is referred to as the \emph{unit} and the
transformation $\epsilon$ is referred to as the \emph{counit}. This is written as:
\[
  (\eta,\epsilon):F\leftadjoint G:A\to B.
\]

Additionally, adjoints are often defined as a one to one correspondence between hom-sets in the
following manner:

Let $F,G$ be functors such that $F:\A \to \B$ and $G:\B \to \A$ such that there is a bijection $\phi$
between the hom-sets:
\[
  \phi: \B(F(X),Y) \to \A(X,G(Y)).
\]
This is the path followed by Mac Lane in \cite{maclan97:categorieswrkmath}, where he proves this
definition is equivalent to Definition~\ref{def:universal_pair} above. Rather than explicitly
defining the bijection, this is often written in the form of an inference:
\[
  \infer{\A(X,G(Y)).}{\B(F(X),Y)}
\]



\begin{example}[Simple slice adjoint]\label{ex:simple-slice-adjoint}
  Consider the  \emph{simple slice} category\cite{blute2009cartesian}, $\B[A]$  defined as:
  \categoryns{Objects of \B.
  }{a map $f_A:C\to D$ in $\B[A]$ is given by the map $f:C\times A \to D$ in \B.
  }{projection -- $\pi_0$.
  }{$C\xrightarrow{f_A}D\xrightarrow{g_A}E$ in $\B[A]$ is given by
    $C\times A \xrightarrow{1\times \Delta}
    C\times A\times A\xrightarrow{f\times 1} D\times A \xrightarrow{g} E$ in \B.}
  Note the definition of simple slice category relies on products in categories, to be introduced
  below in Section~\ref{sub:limits_and_colimits_in_categories}.

  There is an adjunction between $\B[A]$ and $\B$.

  Define $F:\B[A]\to \B$ by $F:C \mapsto C\times A$, $F:(f_A:C\to D) \mapsto (1\times \Delta) (f\times
  1)$.
  Define $G:\B \to \B[A]$ by $G:C\mapsto C$, $G:(f:C\to D) \mapsto (\pi_0 f)_A$. We see immediately
  there is a  correspondence $\phi$ of hom-sets
  \[
    \infer{\B[A](C,G(D))}{\B(F(C), D)}
    \qquad \raisebox{8pt}{$\equiv$}\ \ \infer{\B[A](C,D)}{\B(C\times A, D)}
  \]
  where $\phi:\B(C\times A, D)\to\B[A](C,D)$ is given by $\phi(f) = f_A$.
  Thus $F\leftadjoint G$.
\end{example}

Now, we may consider equivalence of categories.

\begin{definition}\label{def:equivalence_of_categories}
  Given two categories $\A$ and $\B$, we say they are \emph{equivalent} when there exists two
  functors, $F:\A \to \B$ and $G:\B \to \A$ and natural isomorphisms such that $F G \cong I:\A \to
  \A$ and $G F \cong I:\B \to \B$.
\end{definition}


The equivalence functors will also be an adjoint pair, typically they are chosen such that
$F\leftadjoint G$.

\section{Enrichment of categories} % (fold)
\label{sub:enrichement_of_categories}

If $\X$ is a category, then the maps from $A$ to $B$ in $\X$ are denoted $\X(A,B)$.
If $\X(A,B)$ is a set for all objects in $\X$, we say $\X$ is \emph{enriched} in sets. More generally,
categories may be enriched in any monoidal category. For example a category may be
enriched in abelian groups, vector spaces, posets, categories or commutative monoids.

Specific types of enrichment may force a structure on a category. Examples:
\begin{enumerate}
 \item If $\X$ is enriched in sets of cardinality of 0 or 1, then $\X$ is a preorder.
 \item If $\X$ is enriched in pointed sets with the monoid of smash product, then $\X$ has zero morphisms.
% \item If $\X$ is enriched in abelian groups, then $\X$ is a preadditive category (has zero
%   morphisms and has finite products which are the same as the coproduct).
\end{enumerate}


% subsection enrichment_of_categories (end)
\section{Examples of categories} % (fold)
\label{sub:examples_of_categories}
In this section, we will offer a few examples of categories.

\begin{example}
  A group $G$ may be considered as a one-object category $\mathbb{G}$, with object $\{*\}$. The elements of the
  group are the maps between $\{*\}$. As $G$ is a group, it has an identity $e$ and multiplication is
  associative. In $\mathbb{G}$, the identity map is $e$, composition is given
  by the group multiplication and additionally, each map has an inverse. As $\mathbb{G}(\{*\},\{*\})
  = G$, this category is enriched in groups.
\end{example}

Four categories whose objects are sets are \rel, \sets, \Par, and \pinj:

\begin{example}[\rel]\label{ex:category_rel}
\rel is often of interest in quantum programming language semantics:
\category{All sets}{$R:X\to Y$ is a relation: $R\subseteq X\times Y$}{$1_X = \{(x,x) | x \in X\}$}{$RS = \{(x,z) |\,
\exists y. (x,y) \in R$ and $(y,z)\in S\}$}
\end{example}


Note that \rel is enriched in posets (see Example~\ref{ex:preorders-are-categories}), via set
inclusion. \sets and \Par can be viewed as subcategories of \rel, with the same objects, but
restricting the maps allowed --- see below. \Par is also enriched in posets, via the same inclusion
ordering as in \rel.

\begin{example}[\sets]\label{ex:category_sets}
This is the subcategory of \rel which has the same  objects (sets) but only allowing maps which are
total functions, i.e., deterministic relations $f$ where for all $x\in X$, there is a $y$ such
that $(x,y)\in f$ and if $(x,y), (x,y') \in f$, then $y = y'$.
\end{example}

\begin{example}[\Par]\label{ex:category_par}
This is the subcategory of \rel which has the same  objects (sets) but only allowing maps which are
partial functions, i.e., deterministic relations $f$ where if $(x,y), (x,y') \in f$, then $y = y'$.
\end{example}


\begin{example}[\pinj]\label{ex:category_pinj}
Our final example based on sets is one that will be used throughout this thesis. The category \pinj consists of
the partial injective functions over sets. Similarly to \sets and \Par, it is a subcategory
of \rel. The maps $f,g$ (relations in \rel) in \pinj are subject to two conditions:
\begin{align}
   (x,y)\in f\text{ and }(x,y')\in f & \text{ implies } y = y',\label{eq:pinj_relation_is_a_function}\\
   (x,y)\in f\text{ and }(x',y)\in f & \text{ implies } x = x'.\label{eq:pinj_converse_relation_is_a_function}
\end{align}

\end{example}

\begin{example}[\topcat]\label{ex:category_top}
  For a set $S$, $\power{S}$ is the power set of $S$, that is, the set of all subjects of $S$,
  including the empty set and $S$ itself.

  We define $\open{T}$, the \emph{open sets} of $T$, as follows:
  \begin{enumerate}[{(}i{)}]
    \item $\open{T} \subseteq \power{T}$;
    \item $T\in \open{T}$ and $\emptyset \in \open{T}$;
    \item The union of any sets in $\open{T}$ is again in $\open{T}$;
    \item The intersection of finitely many sets in $\open{T}$ is in $\open{T}$.
  \end{enumerate}

  A topological space is defined as a pair $(T,\open{T})$ where $T$ is a set and $\open{T}$ are the
  open sets of $T$. Note that we may choose different open sets for $T$, resulting in different
  topological spaces.

  Maps between two topological spaces $(T,\open{T})$ and $(S,\open{S})$ consist of
  a set map $f:T \to S$ such that the inverse image of any open set in $S$ is an open set in
  $T$. Such maps are referred to as \emph{continuous functions}.

  The identity map is always continuous and composition of continuous functions yields another
  continuous function, hence setting $\topcat_o$ to be the set of all topological spaces and
  $\topcat_m$ to be the continuous maps between them gives us a category.
\end{example}

Our next example shows maps in categories need not always be something normally thought of as a
function or relation.

\begin{example}[Matrix category]\label{ex:matrix_category}
Given a rig $R$ (i.e., a ri\textbf{n}g without \textbf{n}egatives, e.g., the natural numbers), one may
form the category \textsc{Mat}($R$). For example, the category of matrices over natural numbers is:
\category{\nat}{$[r_{i j}]: n \to m$ where $[r_{i j}]$ is an $n \times m$ matrix over \nat}{
$I_n$}{Matrix multiplication}
\end{example}

% \begin{example}[Slice category]\label{ex:slice_category}
%   Given a category $\B$ and an object $A$ in \B, form the \emph{slice category} $\B/A$:
%   \categoryns{Maps of \B whose codomain is $A$, i.e., $f:B\to A$ for some $B$.}{
%     Triples $(f_1,g,f_2):f_1 \to f_2$ such that
%     \[
%       \xymatrix{
%         B_1 \ar[rr]^{g} \ar[dr]_{f_1} && B_2 \ar[dl]^{f_2}\\
%         & A
%       }
%     \]
%     commutes.}{$(f,1_B,f):f \to f$.}{$(f_1,g,f_2)(f_2,h,f_3) = (f_1,g h, f_3)$ which is well
%     defined as $g h f_3 = g f_2 = f_1$.}

%   There is also the concept of the \emph{simple slice} category\cite{blute2009cartesian}, $\B[A]$
%   defined as:
%   \categoryns{Objects of \B.}{$f:C\to D$ is given by maps $f:C\times A \to D$.}{projection --
%     $\pi_0$.}{$C\xrightarrow{f}D\xrightarrow{g}E$ is given by $C\times A \xrightarrow{1\times \Delta}
%     C\times A\times A\xrightarrow{f\times 1} D\times A \xrightarrow{g} E$.}
%   Note the definition of simple slice category relies on products in categories, to be introduced in
%   the next section.
% \end{example}



% subsection examples_of_categories (end)

\section{Limits and colimits in categories} % (fold)
\label{sub:limits_and_colimits_in_categories}

We shall review only a few basic limits/colimits in categories, in order to set up notation and
terminology. First we discuss initial and terminal objects.

\begin{definition}\label{def:initial_object}
  An \emph{initial object} in a category $\B$ is an object which has exactly one map to each other
  object in the category. The dual notion is \emph{terminal object}. Every object in the category
  has exactly one map to the terminal object.
\end{definition}


\begin{lemma}\label{lem:initial_objects_are_unique}
  Suppose $I,J$ are initial objects in $\B$. Then there is a unique isomorphism $i:I \to J$.
\end{lemma}
\begin{proof}
  First, note that by definition there is only one map from $I$ to $I$ --- which must be the
  identity map. As $I$ is initial there is a map $i: I \to J$. As $J$ is initial there is a map
  $j:J \to I$. But this means $i j : I \to I = 1$ and $j i : J \to J = 1$ and hence $i$ is the
  unique isomorphism from $I$ to $J$.
\end{proof}

Dually, we have the corresponding result to Lemma~\ref{lem:initial_objects_are_unique} for terminal
objects --- they are also unique up to a unique isomorphism.

In categories, following the terminology in sets, we normally designate the initial object by $0$
and the terminal object by $1$.
A map \emph{from} the terminal object to another object in the category is often referred to as an
\emph{element} or a \emph{point}.

We now turn to products and coproducts.

\begin{definition}\label{def:categorical_product}
  Let $A,B$ be objects of the category $\B$. Then the object $A \times B$ is a \emph{product} of
  $A$ and $B$ when:
  \begin{itemize}
    \item There exist maps $\pi_0, \pi_1$ with $\pi_0:A\times B \to A$, $\pi_1:A\times B \to B$;
    \item Given an object $C$ with maps $f:C\to A$ and $g:C \to B$ there is a unique map
    $\<f,g\>$ such that the following diagram commutes:
    \[
      \xymatrix@C+5pt@R+15pt{
        &&&A\\
        C\ar[urrr]^{f} \ar[drrr]_{g}\ar@{.>}[rr]^{\<f,g\>} & &A\times B \ar[ur]_{\pi_0}\ar[dr]^{\pi_1}\\
        &&&B.
      }
    \]
  \end{itemize}

\end{definition}

A coproduct is the dual of a product.

\begin{definition}\label{def:categorical_coproduct}
  Let $A,B$ be objects of the category $\B$. Then the object $A + B$ is a \emph{coproduct} of
  $A$ and $B$ when:
  \begin{itemize}
    \item There exist maps $\cpa, \cpb$ with $\cpa:A \to A + B$, $\cpb:B \to A+ B$;
    \item Given an object $C$ with maps $h:A\to C$ and $k:B \to C$ there is a unique map
    $[h,k]$ such that the following diagram commutes:
    \[
      \xymatrix@C+5pt@R+15pt{
        A \ar[drrr]^h \ar[dr]_{\cpa} \\
        &A +B \ar@{.>}[rr]^{[h,k]} && C\\
        B. \ar[urrr]_k \ar[ur]^{\cpb}
      }
    \]
  \end{itemize}
\end{definition}

It is possible for an object to be both a limit and a colimit at the same time:
\begin{definition}\label{def:categorical_zero}
  Given a category $\B$, any object that is both a terminal and initial object is called a
  \emph{zero object}. This object is labelled $\zeroob$.
\end{definition}

Note that any category with a zero object has a special map, $\zeroob_{A,B}$
between any two objects $A,B$ of the category given by: $\zeroob_{A,B} : A \to \mathbf{0} \to B$.

\begin{definition}\label{def:categorical_biproduct}
  In a category \B, with products and coproducts, where:
  \[
    \cp{i}\pi_j = \begin{cases}
      1 & i = j\\
      0 & i \neq j
      \end{cases}
  \]
  and for any two objects, $A,B$, $A\times B$ is the same as $A+B$, then $A\times B$ is
  referred to as the \emph{biproduct} and designated as $A\biproduct B$. A category \cD{} is said to
  have \emph{finite biproducts} when it has a zero object $\mathbf{0}$ and when each pair of objects
  $A,B$ have a biproduct $A\biproduct B$.

  The biproduct is often written as $\+$. This thesis frequently uses $\+$ where it is not a
  biproduct, and thus uses $\biproduct$ for biproduct instead.
\end{definition}

A category with finite biproducts is enriched in commutative monoids. If $f,g:A\to
B$, define $f+g:A\to B$ as $\<1_{A}, 1_{A}\>\, (f\biproduct g)\, [1_{B},1_{B}]$. The unit for
$+$ is $\zeroob_{A,B}$. In this thesis, when discussing a category with finite biproducts,
$\<1, 1\>$ will be designated by $\Delta$ and $[1,1]$ by $\nabla$.

% subsection limits_and_colimits_in_categories (end)


\section{Symmetric Monoidal Categories} % (fold)
\label{sub:categories_with_additional_structure}

\begin{definition}\label{symmetricmonoidalcat}
  A \emph{symmetric monoidal category}\cite{barr:ctcs,maclan97:categorieswrkmath} \cD{} is a
  category equipped with a monoid $\*$ (a bi-functor $\*:\cD \times \cD \to \cD$) together with
  four families of natural isomorphisms:
  \begin{align*}
    u^l_{A}&:I\*A \to A&
    u^r_{A}&:A\*I\to A \\
    a_{A,B,C} &: A\*(B\*C) \to (A\*B)\*C &
    c_{A,B}&:A\*B \to B\* A.
  \end{align*}
  When the objects are identified in other ways, we often write the maps with the tensor as a
  subscript:
  \begin{align*}
    \usl &: I \* A \to A &
    \usr &: A \* I \to A\\
    a_\* &: (A \* B) \* C \to A \* (B \* C) &
    c_\* &: A \* B \to B \* A.
  \end{align*}
  These maps must satisfy the coherence diagrams and equations shown in
  Figures~\ref{fig:SMC_pentagon}, \ref{fig:SMC_unit}, \ref{fig:SMC_commutes},
  \ref{fig:SMC_unit_symmettry} and \ref{fig:SMC_associativity_symmetry}. The essence of the
  coherence diagrams is that any diagram composed solely of the structure isomorphisms will
  commute. The isomorphisms are referred to as the \emph{structure isomorphisms}. $I$ is the
  unit of the monoid. A symmetric monoidal category is called \emph{strict} when each of
  $a_{A,B,C}$, $u^r_{A}$, $u^l_{A}$ and $c_{A,B}$ are identity maps.

  A tensor with these properties is also referred to as a \emph{symmetric tensor}.
\end{definition}

\begin{figure}[!htbp]
\[
  \xymatrix@C+25pt{
    A\*(B\*(C\*D) \ar[r]^{a_{A,B,(C\*D)}} \ar[d]_{1\*a_{B,C,D}}
      & (A\*B)\*(C\*D) \ar[r]^{a_{(A\*B),C,D}}
      & ((A\*B)\*C)\*D \ar[d]^{a_{A,B,C}\*1}\\
    A\*((B\*C)\*D) \ar[rr]_{a_{A,(B\*C),D}}
      && (A\*(B\*C))\*D
  }
\]
\caption{Pentagon diagram for associativity in an SMC.}\label{fig:SMC_pentagon}
\end{figure}
\begin{figure}[!htbp]
\[
  \xymatrix@C+5pt@R+10pt{
    A\*(I\*B) \ar[rr]^{a_{A,I,B}} \ar[dr]_{1\*u^l_B}
      && (A\*I)\*B \ar[dl]^{u^r_A \* 1}\\
      &A\*B
  }
\]
\[\text{ and } u^r_I = u^l_I: I\* I \to I\]
\caption{Unit diagram and equation in an SMC.}\label{fig:SMC_unit}
\end{figure}
\begin{figure}[!htbp]
\[
  \xymatrix@C+5pt@R+10pt{
    A\*B \ar[r]^{c_{A,B}} \ar@{=}[dr]
      & B\*A \ar[d]^{c_{B,A}}\\
      &A\*B
  }
\]
\caption{Symmetry in an SMC.}\label{fig:SMC_commutes}
\end{figure}
\begin{figure}[!htbp]
\[
  \xymatrix@C+5pt@R+10pt{
    A\*I \ar[rr]^{c_{A,I}} \ar[dr]_{u^r_A}
      && I\*A \ar[dl]^{u^l_A}\\
      &A
  }
\]
\caption{Unit symmetry in an SMC.}\label{fig:SMC_unit_symmettry}
\end{figure}
\begin{figure}[!htbp]
\[
  \xymatrix@C+15pt@R+10pt{
    (A\*B)\*C \ar[r]^{c_{(A\*B),C}} \ar[d]_{a^{-1}_{A,B,C}}
      & C\*(A\*B) \ar[d]^{a_{C,A,B}}\\
    A\*(B\*C) \ar[d]_{1\*c_{B,C}}
      & (C\*A)\*B \ar[d]^{c_{C,A}\*1}\\
    A\*(C\*B) \ar[r]^{a_{A,C,B}}
      & C\*(A\*B).
  }
\]
\caption{Associativity symmetry in an SMC.}\label{fig:SMC_associativity_symmetry}
\end{figure}
\begin{example}
  Any category \D with product $\times$ and a terminal object $\top$ is a symmetric monoidal
  category with $\* \definedas \times$ and $I \definedas \top$.
\end{example}
\begin{example}[\pinj has a symmetric tensor]\label{ex:pinj-has-a-smt}
  A symmetric tensor in \pinj is given by the Cartesian product of sets. In detail, this means:
  \begin{align*}
    A \* B &= \{(a,b)| a\in A, b\in B\}\\
    f \* g &= \{((a,c),(b,d)) | (a,b) \in f, (c,d) \in g\}\\
    1 & = \{*\}\text{, a single element set.}
  \end{align*}
  The symmetric monoid isomorphisms are:
    \begin{align*}
      \usl &: \{(*,a)\} \mapsto \{a\}
      &\usr &: \{(a,*)\} \mapsto \{a\}\\
      a_{\*} &: \{((a,b),c)\} \mapsto \{(a,(b,c))\}
      &c_{\*} &: \{(a,b)\} \mapsto \{(b,a)\}.
    \end{align*}
\end{example}
% subsection symmetric monoidal categories (end)
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../phd-thesis"
%%% End:
